---
title: "User Guide: 1 Pre-processing"
subtitle: "Package 'photobiologyFlecks' `r packageVersion('photobiologyFlecks')` "
author: "Pedro J. Aphalo"
date: "`r packageDate('photobiologyFlecks')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide: 1 Pre-processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

At the moment this vignette has the double purpose of keeping track of my
progress and testing the performance of the functions in the package. My aim
is to analyse a large set of high frequency time series measurements of
sunlight at seven different wavebands done with broadband sensors over one
summer.

The data set will be large with seven variables and a few hundred millions of
time points. During daytime measurements were acquired in bursts of 18000 time
points in 15 min every 30 min.

The effect of clouds is intermittent and varies in intensity gradually. Using
quantiles, mean, MAD and variance on the irradiance, and on the running
differences is one possible approach. However, a recent additional approach
is based on detecting change points in the running differences as a basis to
quantify the duration and amplitude of cloud/shade flecks.

The data are in large text files saved with the software PC400 from Campbell
Scientific, the supplier of the data logger used to collect the data set.

The first step is to read-in the data and split it into "chunks" of data from
18000 consecutive observations at the gaps between the "bursts" of observations,
discarding any incomplete sets and any data measured at a lower frequency. At
this stage running differences for time stamps need to be computed, and running
differences for the measured variables can also be added at the same time.

The smaller values in the running differences are most likely the result of
measurement "noise" and could be "cleaned" before the next steps. The
differences can be used directly or discretized based on their sign.

## Current state

The reading of the data from the 3.7GB text file into a tibble is the slowest
step. Other computations are fast even with the functions are coded fully in R.

## Set up

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(photobiology)
library(photobiologyInOut)
library(photobiologyFlecks)
library(ggplot2)
```

## The data

We use data included in the package as a data frame. These data were originally saved as a CSV file by Campbell Scientific's PC400 program. The data were acquired with a CR6 data logger.

```{r}
head(three_chunks.tb)
```

```{r}
plot(three_chunks.tb, cex = 0.1)
```

## Split data into chunks

The time series data of sunlight have been measured in 15-min-long bursts separated by 15 min with no data acquisition. Data from each burst will be analysed separately. For this the data frame has to be split at the gaps between bursts.

```{r}
# keep all qty columns, add differences
chunks.ls <-
  split_chunks(three_chunks.tb,
               time.name = "time",
               qty.name = NULL,
               chunk.min.time = 0.051,
               chunk.min.rows = 1.8e4,
               verbose = FALSE)
```

Some checks

```{r}
length(chunks.ls) == length(unique(names(chunks.ls)))
nchunks_found <- length(chunks.ls)
```

```{r}
str(chunks.ls, give.attr = FALSE)
```
```{r}
plot(Q_PAR ~ time, data = chunks.ls[[1]], type = "l", lwd = 0.5)
plot(Q_PAR.diff ~ time, data = chunks.ls[[1]], type = "l", lwd = 0.5)
```

## Summaries

### Statistics

Something seems wrong as I expect a 1 h time shift to be needed!

```{r}
statistics.tb <- 
  summarize_chunks(chunks.ls, 
                   add.times = TRUE, 
                   tz = "EET",
                   time.shift = 0,
                   add.solar.times = TRUE,
                   geocode = data.frame(lon = 25.019212, lat = 60.226805),
                   verbose = FALSE)
```

```{r}
colnames(statistics.tb)
length(unique(statistics.tb[["chunk"]])) == nchunks_found
length(unique(statistics.tb[["time"]])) == nchunks_found
```


```{r}
statistics.tb
```

## Locating change points

### Using `diff()`

Running differences after slight denoising can be used to detect changes in the sign of the local slope, the peaks and valleys of a time series. In other words, the flecks are identified based on local maxima and minima.

We use `relative.threshold = 0.20` $\mu mol\,m^{-2}\,s^{-2}$ as the minimum rate of change in irradiance per second to be considered relevant and not discarded as noise.

```{r}
  chunks_dn.ls <- 
    denoise_chunks(chunks.ls, 
                   time.name = "time",
                   relative.threshold = 0.0067, 
                   add.signs = TRUE)
```

```{r}
plot(Q_PAR.diff ~ time, data = chunks_dn.ls[[1]], type = "l", lwd = 0.5)
```

In a continuous time series without noise, the derivative is equal to zero at the point where the slope changes sign. In a time series measured at discrete time points, a change in sign must be used as well as a zero value to locate maxima and minima, as the change in slope is most likely to land in-between two time points.

Once de-noised, changes in slope can be detected using `sign()` to discretize the data into values indicating decrease (-1) , no-change (0) and increase (+1).

Considering each of the three bursts of PAR measurements, we find many time intervals with a detectable change in irradiance.

```{r}
# time steps with no change
sum(abs(chunks_dn.ls[[1]]$Q_PAR.diff) < 0.01, na.rm = TRUE)
# increasing time steps
sum(chunks_dn.ls[[1]]$Q_PAR.diff > 0.01, na.rm = TRUE)
# decreasing time steps
sum(chunks_dn.ls[[1]]$Q_PAR.diff > 0.01, na.rm = TRUE)
```

```{r}
# time steps with no change
sum(abs(chunks_dn.ls[[2]]$Q_PAR.diff) < 0.01, na.rm = TRUE)
# increasing time steps
sum(chunks_dn.ls[[2]]$Q_PAR.diff > 0.01, na.rm = TRUE)
# decreasing time steps
sum(chunks_dn.ls[[2]]$Q_PAR.diff > 0.01, na.rm = TRUE)
```
```{r}
# time steps with no change
sum(abs(chunks_dn.ls[[3]]$Q_PAR.diff) < 0.01, na.rm = TRUE)
# increasing time steps
sum(chunks_dn.ls[[3]]$Q_PAR.diff > 0.01, na.rm = TRUE)
# decreasing time steps
sum(chunks_dn.ls[[3]]$Q_PAR.diff > 0.01, na.rm = TRUE)
```

Subsequently, changes in slope can be located using `rle()`. In this example all differences smaller than 1 % of the mean irradiance per second were discarded as
noise.

```{r}
sign_changes.ls <- list()
for (i in names(chunks_dn.ls)) {
  sign_changes.ls[[i]] <- 
    list(rle = rle(chunks_dn.ls[[i]][["Q_PAR.sign"]]))
}
length(sign_changes.ls)
```
Check that sum of lengths equal length of burst.

```{r}
sum(sign_changes.ls[[1]][["rle"]][["lengths"]])
```

```{r}
sum(sign_changes.ls[[2]][["rle"]][["lengths"]])
```

```{r}
sum(sign_changes.ls[[3]][["rle"]][["lengths"]])
```

```{r}
sign_changes.ls
```
### Durand's et al. algorithm

```{r}
# keep all qty columns, do not add differences
chunks.ls <-
  split_chunks(three_chunks.tb,
               time.name = "time",
               qty.name = NULL,
               chunk.min.time = 0.051,
               chunk.min.rows = 1.8e4,
               add.diffs = FALSE,
               verbose = FALSE)
```

```{r}
chunk <- chunks.ls[[1]]
chunk$tod <- as_tod(chunk$time, unit.out = "seconds")
chunk$chuck.time <- chunk$tod - chunk$tod[1]

Z <- find_zeros(time = chunk$chuck.time,
                var = chunk$Q_PAR)
dfS <- find_flecks(time = chunk$chuck.time,
                   var = chunk$Q_PAR,
                   zeroes = Z,
                   minTime = 0, 
                   minAmp = 5, 
                   minPdiff = 0.05,
                   asymmetry = 1/4,
                   shadeflecks = FALSE,
                   asmMethod = "max",
                   verbose = FALSE)
plot_ts_fleck(time = chunk$chuck.time,
              var = chunk$Q_PAR,
              zeroes = Z,
              fleck.data = dfS)
```

### Using a reference

Instead of detecting sunflecks as a dynamic temporal process, or local changes in slope, especially in the case of cloud flecks an alternative is to use the estimated clear-sky irradiance as reference and to quantify the size of the deviations, both increases and decreases, in many cases using a fixed threshold. An alternative to using a clear sky simulation is to use a smoothed version of the same data as a reference, a moving median, or a spline-based smoother. However, each of these and the temporal-process approach described above are fundamentally different and describe different properties of the data.

Both observed and reference irradiances need to be available at the same time points. This in some cases can make it necessary to use interpolation to obtain suitable reference data. After this, a simple difference at each time point can be computed and compared against a threshold. As the reference varies in time, the threshold used can be expressed either as an absolute difference or as a relative difference, thus, giving two different quantifications.

### Using a quantile smoother

Quantile regression, either of a linear model or a smmothing spline, differs from OLS regression and other common approaches in that it can flexibly be used to find a fitted line that describes a specific quantile. It can be imagined as a combination of a boxplot and a smoother. This approach lets us estimate local deviations from an internal reference different from the local mean or median, say, for example, to identify when irradiance is less than that in the 10% locally highest irradiances.





 For vegetation shade a further possibility is to use a quantile regression directly fitted for a target shade threshold.

